Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks

### 2019-04-02 doc/paper/paper.tex
3 body problem paper written

\usepackage{arxiv}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{lipsum}
\usepackage{caption}
\usepackage{float}
\usepackage{graphicx}

@@

When forecasting, the goal is to have accurate predictions. However, 
a chaotic system's sensitivity to initial conditions implies that a small 
perturbation as a result of error will produce incorrect solutions.
Further, given a periodic
solution, we hope that a small perturbation, from error in measurement or from 
floating-point errors in numerical calculations, doesn't affect the solution or 
at least produces a quasi-periodic solution close to the periodic one. 
(A solution or trajectory $F$ 
is quasi-periodic if and only if for all $t$ and for some $\tau$, $F(t+\tau)$ 
is arbitrarily close to $F(t)$ \cite{lorenz1963deterministic}.)

An example of this is the trajectories of the sun and planets in our solar 
system. Newton's equations can be used to create a system of equations to model 
the bodies of mass (known as the n-body problem). However, with the existence of
other bodies of mass from other solar systems within the universe, we will have 
small perturbations introduced into the system. If the system is chaotic, then 
over time, the actual trajectories will diverge from our model's 
\cite{oestreicher2007history}.

@@

In terms of forecasting, it's useful to know how chaotic a dynamical system is. 
To measure this, we will measure the rate at which nearby trajectories 
diverge. Constants describing the divergence are known as the Lyapunov 
Characteristic Exponents (LCEs) \cite{sandri1996numerical}. There is an LCE for 
each dimension, and since they all represent exponential growth. The error 
growing at the rate of the largest LCE will dominate over the others.

@@

Three dynamical systems will be used to demonstrate the predictability of 
chaotic systems: the Lorenz system, Poincaré's 3-body 
problem solution, and the double pendulum system.
The Poincaré system is the one of the first (possibly the first), and Lorenz
is a classical example. The double pendulum is commonly referred to as chaotic, 
and I will show this later.

@@

There are however two issues when integrating this problem. As objects approach each
other, the distance between them 
approaches zero. If we have a collision, then this approaches infinity. There is no 
collision detection in this model as we assume falsely that each body has a radius of zero,
so this case is not handled. Instead, we will get a 
divide by zero error, and the model cannot be calculated further. Further, near collisions 
are also an issue. Detailed in \cite{chambers1999hybrid}, with the distance between bodies 
approaching zero, their potential energy becomes significantly smaller than other values, 
such as those describing kinetic energy. This 
produces a stiff system of equations, and introduces error. Since energy is conserved in
this system (as it is a Hamiltonian system), the error introduced represents energy being 
created. This causes the singularity property seen in Figure \ref{fig:singularity}

\begin{figure}[h]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision.png}
        \caption{The restricted 3 body problem with collision}
        \label{fig:r3b_collision}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
        \caption{The resultant divergence of the body}
        \label{fig:r3b_collision_whole}
    \end{subfigure}
    \caption{The Double Pendulum}
    \label{fig:doub_pend}
\end{figure}

While it is possible to solve this issue with symplectic integrators \cite{chambers1999hybrid},

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=.5\linewidth]{images/example_doub_pend.png}
    \caption{The Double Pendulum}
    \label{fig:doub_pend}
\end{figure}

@@

>>>>

\usepackage{arxiv}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{lipsum}
\usepackage{caption}
\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}

@@

When forecasting, the goal is to have accurate predictions. However, 
a chaotic system's sensitivity to initial conditions implies that a small 
perturbation as a result of error will produce incorrect solutions.
Further, given a periodic
solution, the hope is that a small perturbation, from error in measurement or from 
floating-point errors in numerical calculations, doesn't affect the solution or 
at least produces a quasi-periodic solution close to the periodic one. 
(A solution or trajectory $F$ 
is quasi-periodic if and only if for all $t$ and for some $\tau$, $F(t+\tau)$ 
is arbitrarily close to $F(t)$ \cite{lorenz1963deterministic}.)

An example of this is the trajectories of the sun and planets in our solar 
system. Newton's equations can be used to create a system of equations to model 
the bodies of mass (known as the n-body problem). However, with the existence of
other bodies of mass from other solar systems within the universe, 
small perturbations are introduced into the system. If the system is chaotic, then 
over time, the actual trajectories will diverge from our model's 
\cite{oestreicher2007history}.

@@

In terms of forecasting, it's useful to know how chaotic a dynamical system is. 
Chaos will be measured by the exponential rate at which nearby trajectories 
diverge. Constants describing the divergence are known as the Lyapunov 
Characteristic Exponents (LCEs) \cite{sandri1996numerical}. There is an LCE for 
each dimension, and since they all represent exponential growth. The error 
growing at the rate of the largest LCE will dominate over the others.

@@

Three dynamical systems will be used to demonstrate the predictability of 
chaotic systems: the Lorenz system, Poincaré's 3-body 
problem solution, and the double pendulum system.
The Poincaré system is the one of the first, and the Lorenz system
is a classic example. The double pendulum is commonly referred to as chaotic, 
and I will show this later. Both Poincaré's restricted 3 body problem solution
and 
the double pendulum are Hamiltonian systems representing physical systems
within our universe.

@@

There are however two issues when integrating this problem. As objects approach each
other, the distance between them 
approaches zero. If collision occurs, then this approaches infinity. There is no 
collision detection in this model, and each body has a radius of zero,
so this case is not handled. Instead, a divide by zero error occurs,
and the model cannot be calculated further.

Further, near collisions 
are also an issue. With the distance between bodies 
approaching zero, their potential energy becomes significantly smaller than other values, 
such as those describing kinetic energy \cite{chambers1999hybrid}. This 
produces a stiff system of equations, and introduces error. Since energy is conserved in
this Hamiltonian system, the error introduced represents energy being 
created. This can cause the singularity property seen in Figure \ref{fig:singularity}.
This problem can be solved by using a symplectic integrator \cite{chambers1999hybrid}.

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision.png}
        \caption{The restricted 3 body problem with collision}
        \label{fig:r3b_collision}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
        \caption{The resultant divergence of the body}
        \label{fig:r3b_collision_whole}
    \end{subfigure}
    \caption{The Restricted 3 Body Problem Error from Collision}
    \label{fig:singularity}
\end{figure}

While it is possible to use a symplectic integrator, it is not necessary. Under the right 
initial conditions, the third body will diverge regardless. The simple divergent spiral
isn't chaotic, so instead orbital and collision less initial conditions will be used.
Without divergence and without collisions, the symplectic integrator is unnecessary.

@@

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=.5\linewidth]{images/example_doub_pend.png}
    \caption{The Double Pendulum}
    \label{fig:doub_pend}
\end{figure}
